---
title: "Session 10 lab exercise: Repeated Measures and Longitudinal Analysis II"
author: "Levi Waldron"
institute: "CUNY SPH Biostatistics 2"
clean: false
output:
  html_document:
    toc: yes
    df_print: paged
    theme: lumen
    number_sections: yes
  md_document:
    preserve_yaml: false
always_allow_html: true
---

**Learning objectives**

1. Gain an intuitive understanding of ICC through simulated data
2. Simulate correlated grouped data
3. Use a heatmap and spaghetti plot to visualize correlated grouped data
4. Fit random and mixed-effects models to correlated grouped data
5. Calculate ICC from a random or mixed-effects model
6. Fit a population average model, aka marginal model, using GEE

**Exercises**

1. Simulation of correlated grouped data
2. Create a heatmap of simulated data to visualize the group effect
3. Create a spaghetti plot of the simulated data to visualize the group effect
4. Fit a random effects model with no covariates and a random intercept. Does it recover the group and residual variances you simulated?
5. Estimate ICC from the model above. Is it what you expected from the group and residual variances you simulated?
6. Estimate ICC simply by calculating the correlation between fecfat1 and fecfat2. Is it similar to the estimate above?
7. Load and do basic cleaning of the Georgia Birthweights dataset.
8. Make a boxplot and spaghetti plot for the Georgia Birthweights dataset
9. Test the null hypotheses that baseline birth weights do not vary by mother
10. Create QQ plots of residuals and random intercepts for this model.
11. Test the null hypotheses that the effect of birth order not modified by mother’s age at first birth or weight of first infant.

# Simulation of correlated grouped data

Simulate a dataset with two fecal fat measurements on each of `n` study subjects, where the measurement is the sum of a subject mean plus random measurement error. Subject means are distributed $N(10, \sigma_{subj})$ and measurement errors are distributed $N(0, \sigma_{resid})$. Start with the following values:
```{r}
sigma_subj <- sqrt(3)
sigma_resid <- 1
n <- 100
```

```{r}
library(tidyverse)
df <- tibble(subj_mean = rnorm(n, mean = 10, sd = sigma_subj)) %>%
  mutate(id=factor(1:n)) %>%
  mutate(fecfat1 = rnorm(n, mean = 0, sd = sigma_resid) + subj_mean) %>%
  mutate(fecfat2 = rnorm(n, mean = 0, sd = sigma_resid) + subj_mean)
```

# Create a heatmap of simulated data to visualize the group effect

```{r}
library(pheatmap)
pheatmap(select(df, fecfat1:fecfat2))
```

Also play with the values of `sigma_subj` and `sigma_resid` to see what effect this has on the heatmap.

# Create a spaghetti plot of the simulated data to visualize the group effect

Also play with the values of `sigma_subj` and `sigma_resid` to see what effect this has on the spaghetti plot

```{r}
library(ggplot2)
df2 <- pivot_longer(df, cols = starts_with("fecfat"))
ggplot(df2, aes(x=name, y=value, group = id)) + geom_line(alpha=0.25)
```

# Fit a random effects model with no covariates and a random intercept. Does it recover the group and residual variances you simulated?

```{r}
library(nlme)
fit <- lme(value ~ 1, random = ~1|id, data=df2)
summary(fit)
intervals(fit)
```

# Estimate ICC from the model above. Is it what you expected from the group and residual variances you simulated?

Recall ICC for subject $i$, measurements $j$ and $k$:
\begin{equation*}
\begin{aligned}
ICC & = corr(x_{ij}, x_{ik}) \\
    & = \frac{\sigma_{subj}^2}{\sigma_{subj}^2 + \sigma_{\epsilon}^2} \\
    & = \frac{\tau_{00}^2}{\tau_{00}^2 + \sigma_\epsilon^2}
\end{aligned}
\end{equation*}

```{r}
VarCorr(fit)
```

```{r}
varests <- as.numeric(VarCorr(fit)[1:2])
varests[1] / sum(varests)
```

# Estimate ICC simply by calculating the correlation between `fecfat1` and `fecfat2`. Is it similar to the estimate above?

```{r}
with(df, cor(fecfat1, fecfat2))
```

# Load and do basic cleaning of the Georgia Birthweights dataset. 

a. Fix NA values for `momage`
b. Create a categorical age variable with cut at age <18 vs >=18
c. Convert `momid` to a factor
d. Recode the low birthweight variable to a factor, with "0" to "normal" as the reference group and "1" to "low".

```{r, message=FALSE}
library(tidyverse)
ga <- read_csv("gababies.csv") %>%
  mutate(momage = na_if(momage, 99)) %>%
  mutate(agebin = cut(initage, breaks = c(0, 17, 100))) %>%
  mutate(momid = factor(momid)) %>%
  mutate(lowbrth = recode_factor(lowbrth, `0` = "normal", `1` = "low"))
```

# Make a boxplot and spaghetti plot for the Georgia Birthweights dataset

```{r boxplot, fig.cap="Boxplot of Georgia Birthweights dataset"}
ggplot(ga, aes(x = birthord, group = birthord, y = bweight)) +
  geom_boxplot(notch = FALSE, outlier.shape = NA) +
  geom_jitter(width=0.2, col = "black", alpha = 0.25) +
  theme_grey(base_size = 16) + 
  labs(subtitle = "Georgia birthweight dataset") +
  xlab("Birth order") + ylab("Birth Weight (g)")
```

```{r spaghetti, fig.cap="Spaghetti plot of Georgia Birthweights dataset"}
ggplot(ga, aes(x = birthord, y = bweight, group = momid)) +
  geom_line(alpha=0.5) +
  theme_grey(base_size = 16) + 
  theme(legend.position = "none") +
  labs(subtitle = "Georgia birthweight dataset") +
  xlab("Birth order") + ylab("Birth Weight (g)")
```

# Test the null hypotheses that baseline birth weights do not vary by mother

```{r}
library(nlme)
gafit1 <- lme(bweight ~ birthord, data=ga, 
    random=~1|momid)
```

```{r}
intervals(gafit1)
```

# Create QQ plots of residuals and random intercepts for this model.

```{r}
qqnorm(residuals(gafit1, type = "pearson"), main = "QQ plot residuals")
qqline(residuals(gafit1, type = "pearson"))
```

```{r}
head(ranef(gafit1))
qqnorm(ranef(gafit1)[, 1], main = "QQ plot random intercepts")
qqline(ranef(gafit1)[, 1])
```

# Test the null hypotheses that the effect of birth order not modified by mother's age at first birth or weight of first infant.


```{r}
gafit2 <- lme(bweight ~ birthord*agebin, data=ga, 
    random=~1|momid)
summary(gafit2)
intervals(gafit2)
```

```{r}
gafit2 <- lme(bweight ~ birthord*initwght, data=ga, 
    random=~1|momid)
summary(gafit2)
intervals(gafit2)
```
